## Statistics for Laboratory Scientists (140.615)
## Lab 3                                      

### Binomial distribution

```{r}
dbinom(0:7,7,0.2)
dbinom(0:7,7,0.6)

dbinom(0:4,7,0.2)
dbinom(0:4,7,0.6)

pbinom(0:7,7,0.2)
pbinom(0:7,7,0.6)

sum(dbinom(0:4,7,0.2))
sum(dbinom(0:4,7,0.6))

pbinom(0:7,7,0.2)[5]
pbinom(0:7,7,0.6)[5]
```

#### Plotting the distribution

```{r}
barplot(dbinom(0:7,7,0.2),main="Binomial(n=7,p=0.1)",ylim=c(0,0.45),names=0:7,col=grey(0.9))
barplot(dbinom(0:7,7,0.6),main="Binomial(n=7,p=0.6)",ylim=c(0,0.45),names=0:7,col=grey(0.3))
```

#### Two panels per plot

```{r}
par(mfrow=c(1,2))
barplot(dbinom(0:7,7,0.2),main="Binomial(n=7,p=0.1)",ylim=c(0,0.45),names=0:7,col=grey(0.9))
barplot(dbinom(0:7,7,0.6),main="Binomial(n=7,p=0.6)",ylim=c(0,0.45),names=0:7,col=grey(0.3))

par(mfrow=c(2,1))
barplot(dbinom(0:7,7,0.2),main="Binomial(n=7,p=0.1)",ylim=c(0,0.45),names=0:7,col=grey(0.9))
barplot(dbinom(0:7,7,0.6),main="Binomial(n=7,p=0.6)",ylim=c(0,0.45),names=0:7,col=grey(0.3))
```

#### Side by side

```{r}
z <- rbind(dbinom(0:7,7,0.2),dbinom(0:7,7,0.6))
par(mfrow=c(1,1))
barplot(z,beside=T,names=0:7,col=grey(c(0.9,0.3)),space=c(0,0.5),ylim=c(0,0.45))
legend(0,0.45,c("Binomial(n=7,p=0.1)","Binomial(n=7,p=0.6)"),
       pch=22,pt.bg=grey(c(0.9,0.3)),pt.cex=2,bty="n")
```

Remember: you can look at the help file using the '?' character. Type '?legend' to see all options. Type 'example(legend)' to see examples.


### for() loops

```{r}
a <- 1:10
b <- 2:11
d <- rep(NA,10)
d

# multiply two vectors
for(i in 1:10){
  d[i]<-a[i]*b[i]
}
d

# compute the inner product
s<-0
for(i in 1:10){
  s<-s+d[i]
}
s
```

Note that for loops can often be avoided by using vectorized operations. Better ways to do the above are:

```{r}
# multiply two vectors
d <- a*b
d

# compute the inner product
s <- sum(a*b)
s
```

Note that for() can iterate over any sequence:

```{r}
for(j in c(3,1,4,1,5,9,2,7)) print(j)
for(j in LETTERS[1:10]) print(j)
```


### Non-Binomial vs Binomial

This is an example from class, using the for() loop. You will also need the sample() and ifelse() functions.

```{r}
?sample
sample(1:10)
sample(1:10,replace=TRUE)
sample(1:10,repl=T)

f1 <- c("A","a")
f2 <- sample(f1,2,repl=T)
f2

?ifelse
ifelse(3+5<9,"Yes","No")
ifelse(3+8<9,"Yes","No")

f2
ifelse("A"%in%f2,1,0)
```

Consider Mendel's pea experiments.

Purple or white flowers, purple dominant to white.

F0 genotypes are PP and ww, F1 genotypes are Pw.

Pick a random F2.

Self it and acquire 10 progeny.

The number of progeny with purple flowers is not binomial.

```{r}
nit <- 10000
npurple <- rep(0,nit)
f1 <- c("A","a")
for(k in 1:nit){
  f2 <- sample(f1,2,repl=T)
  purple <- rep(0,10)
  for(j in 1:10){
    f3 <- sample(f2,2,repl=T)
    purple[j] <- ifelse("A"%in%f3,1,0)
  }
  npurple[k] <- sum(purple)
}

table(npurple)
table(factor(npurple,levels=0:10))

par(mfrow=c(1,1))
barplot(table(factor(npurple,levels=0:10)))
```

Pick 10 random F2's.

Self each and take a child from each.

The number of progeny with purple flowers is binomial.

```{r}
nit <- 10000
npurple2 <- rep(0,nit)
f1 <- c("A","a")
for(k in 1:nit){
  purple <- rep(0,10)
  for(j in 1:10){
    f2 <- sample(f1,2,repl=T)
    f3 <- sample(f2,2,repl=T)
    purple[j] <- ifelse("A"%in%f3,1,0)
  }
  npurple2[k] <- sum(purple)
}

table(npurple2)
table(factor(npurple2,levels=0:10))
barplot(table(factor(npurple2,levels=0:10)))

par(mfrow=c(1,2))
barplot(table(factor(npurple,levels=0:10)),ylim=c(0,3000),main="Non - Binomial")
barplot(table(factor(npurple2,levels=0:10)),ylim=c(0,3000),main="Binomial")
```

### Poisson distribution

```{r}
dpois(0:10,5)
sum(dpois(0:10,5))
ppois(0:10,5)
ppois(10,5)

dpois(0:15,5)
ppois(0:15,5)
1-ppois(0:15,5)
ppois(0:15,5,lower=FALSE)
```

#### Poisson approximation for the Binomial

```{r}
n=50000
p=1/100000
lambda=n*p

dbinom(0:5,n,p)
barplot(dbinom(0:5,n,p),main="Binomial",ylim=c(0,0.65),names=0:5,col=grey(0.9))

dpois(0:5,lambda)
barplot(dpois(0:5,lambda),main="Poisson",ylim=c(0,0.65),names=0:5,col=grey(0.3))

z <- rbind(dbinom(0:5,n,p),dpois(0:5,lambda))
par(mfrow=c(1,1))
barplot(z,beside=T,names=0:5,col=grey(c(0.9,0.3)),space=c(0,0.5),ylim=c(0,0.65))
legend(4,0.65,c("Binomial","Poisson"),pch=22,pt.bg=grey(c(0.9,0.3)),pt.cex=2,bty="n")

dbinom(0:5,n,p)-dpois(0:5,lambda)
```

## End of Lab 3
